---
title: 'Reproducible Research: Peer Assessment 1'
---
COURSERA REVIEWER: please see figures folder if the plots do not display in your browser. Thanks!
Connect with me on Linkedin: https://www.linkedin.com/in/hadrien-dykiel-29a78432

#Set up R environment
Set working directory and load requried packages.
```{r}
library(ggplot2)
library(plyr)
library(mice) #used to impute missing values
library(chron)
library(cowplot)#needed to graph multiple plots

```


## Loading and preprocessing the data
Save the CSV file to a data frame. Remove NAs. Summarize data set for later use.
```{r}
df <- read.csv('Data/activity.csv') #Load Data
df2 <- df[!is.na(df$steps), ] #remove NAs
df2_summarized <- ddply(df2, c("date"), summarize, sumSteps = sum(steps)) #summarize

```



## What is mean total number of steps taken per day?
Make a histogram of the total number of steps taken each day:
```{r}
hist <- qplot(df2_summarized$sumSteps,
      xlab='Total Steps Each Day',
      ylab='Frequency')

#save to figures folder
png('figures/totalStepsPerDay.png', width=1000)
hist
dev.off()

#print plot
hist
```

Plot total number of steps per date (out of curiousity, not part of assignment)
```{r}
#plot results
ggplot(data=df2_summarized, mapping = aes(x=date, y=sumSteps)) +
  geom_bar(stat="identity", fill = 'blue') +
  theme(axis.text.x = element_text(angle = 90)) + #rotate x axis labels
  ggtitle('Steps Per Day') +
  theme(plot.title = element_text(face='bold', size=16))



#Calculate mean & median
mean <- mean(df2_summarized$sumSteps, na.rm = TRUE)
median <- median(df2_summarized$sumSteps, na.rm = TRUE)

```
The mean number of steps is `r mean` and the median is `r median`.


## What is the average daily activity pattern?
```{r}
timeSeries <- ggplot(data=df2, aes(interval, steps)) +
  stat_summary(fun.y = 'mean', geom = 'line') +
  ggtitle('Avg Steps Per 5 Minute Interval') +
  theme(plot.title = element_text(face='bold', size=16))

timeSeries #print plot

#save to figures folder
png('figures/activityPattern.png', width=1000)
timeSeries
dev.off()



avg_steps_by_interval <- ddply(df2, c("interval"), summarize, avgSteps = sum(steps)) #summarize

#Print busiest time
busiestInterval <- (avg_steps_by_interval$interval[avg_steps_by_interval$avgSteps == max(avg_steps_by_interval$avgSteps, na.rm = T)])

```

The busiest interval is `r busiestInterval`.

## Impute missing values and re-plot results
We'll use the mice package that imputes missing values with plausible values "drawn from a distribution specifically designed for each missing datapoint"(http://www.r-bloggers.com/imputing-missing-data-with-r-mice-package/).
```{r}
sum(is.na(df$steps)) #calculate sum of NAs
tempData <- mice(df, seed = 500) #impute missing values
df3 <- complete(tempData, 1) 
df3_summarized <- ddply(df3, c("date"), summarize, sumSteps = sum(steps)) #summarize

hist3 <- qplot(df3_summarized$sumSteps,
      xlab='Total Steps Each Day (Imputed)',
      ylab='Frequency')

#save to figures folder
png('figures/totalStepsPerDayImputed.png', width=1000)
hist3
dev.off()

#print plot
hist3

#Plot steps per day
ggplot(data=df3_summarized, mapping = aes(x=date, y=sumSteps)) +
  geom_bar(stat="identity", fill = 'blue') +
  theme(axis.text.x = element_text(angle = 90)) + #rotate x axis labels
  ggtitle('Steps Per Day (Imputed)') +
  theme(plot.title = element_text(face='bold', size=16))


ggplot(data=df3, aes(interval, steps)) +
  stat_summary(fun.y = 'mean', geom = 'line') +
  ggtitle('Avg Steps Per 5 Minute Interval (Imputed)') +
  theme(plot.title = element_text(face='bold', size=16)) 

#Calculate mean & median
mean_imputed <- mean(df3_summarized$sumSteps, na.rm = TRUE)
median_imputed <- median(df3_summarized$sumSteps, na.rm = TRUE)


```
After imputation, the new mean and median are ` r mean_imputed` and `r median_imputed`, respectively.


## Are there differences in activity patterns between weekdays and weekends?
```{r}
df3$weekend <- is.weekend(df3$date)
for (i in 1:nrow(df3)) {

  if (df3$weekend[i] == TRUE) {
    df3$dayType[i] <- 'weekend'
  } else {
    df3$dayType[i] <- 'weekday'
  }
}

plot_weekend <- ggplot(data=subset(df3, dayType == 'weekend'), aes(interval, steps)) +
  stat_summary(fun.y = 'mean', geom = 'line') +
  ggtitle('Weekend Avg Steps Per 5 Minute Interval (Imputed)') +
  theme(plot.title = element_text(face='bold', size=16)) 

plot_weekday <- ggplot(data=subset(df3, dayType == 'weekday'), aes(interval, steps)) +
  stat_summary(fun.y = 'mean', geom = 'line') +
  ggtitle('Weekday Avg Steps Per 5 Minute Interval (Imputed)') +
  theme(plot.title = element_text(face='bold', size=16)) 

#Save plot to figures folder
png('figures/final_figure.png')
plot_grid(plot_weekday, plot_weekend, ncol = 1, nrow=2)
dev.off()

#Print plot
plot_grid(plot_weekday, plot_weekend, ncol = 1, nrow=2)
```


